---
title: "MAT442 - Project01"
author: "Titus S Ammerman"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document:
    df_print: kable
---

```{r setup, include=FALSE}
# don't forget to name and number tables
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```


> For centuries, people looked at the full moon with some trepidation. From
stories of werewolves coming out, to more crime sprees, the full moon has
gotten the blame. Some researchers (I will send you the article later this
week) in the early 1970s set out to actually study whether there is a “fullmoon”
effect on the mental health of people. The researchers collected
admissions data for the emergency room at a mental health hospital for 12
months. They separated data into rates before the full moon (mean number
of patients seen 4 to 13 days before the full moon), during the full moon
(the number of patients seen on the full moon day), and after the full moon
(mean number of patients seen 4 to 13 days after the full moon). They also
kept track of which month the data came from since there was likely to be
a relationship between admissions and the season of the year. The data can
be found on D2L in the file MentalHealth.csv (also attached in the D2L
Submission folder). Analyze the data to answer the researchers’ question.

# Abstract

## Purpose

## Background

 - Sample:
 - Population of interest:
 - Relevant variables: var1(category), var2(category)
 
# Exploratory Data Analysis

```{r}
# load dataset...

MentalHealth <- read_csv("MentalHealth.csv", show_col_types = FALSE)

# defining the month order
month_order <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                 "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
# defining the lunar order
moon_order <- c("Before", "During", "After")

# converting month and moon columns to factors with specified levels
MentalHealth <- MentalHealth %>%
  mutate(Moon = factor(Moon, levels = moon_order),
         Month = factor(Month, levels = month_order))

# summary statistics for overall sample

MentalHealth

```

Now we will break this sample down according to `groups`, starting with...

```{r}
# grouping by lunar phase
ggplot(MentalHealth, aes(x = Moon, y = Admission, fill = Moon)) +
  geom_boxplot() +
  theme_minimal()

# grouping by month
ggplot(MentalHealth, aes(x = Month, y = Admission, fill = Month)) +
  geom_boxplot() +
  theme_minimal()
```

We will crease a `season` variable to observe trends that may be less apparent when viewing
admissions on a monthly basis:

```{r}
MentalHealth <- MentalHealth %>%
  mutate(Season = case_when(
    Month %in% c("Dec", "Jan", "Feb") ~ "Winter",
    Month %in% c("Mar", "Apr", "May") ~ "Spring",
    Month %in% c("Jun", "Jul", "Aug") ~ "Summer",
    Month %in% c("Sep", "Oct", "Nov") ~ "Fall"))

# grouping by season
ggplot(MentalHealth, aes(x = Season, y = Admission, fill = Season)) +
  geom_boxplot() +
  theme_minimal()
```
```{r}
# Sum admissions by Month and Moon Phase
contingency_table1 <- xtabs(Admission ~ Month + Moon, data = MentalHealth)
contingency_table2 <- xtabs(Admission ~ Season + Moon, data = MentalHealth)

# Print the contingency table
print(contingency_table1)
print(contingency_table2)

```
# Model Selection

## Model Assumptions

# Conclusion

## Limitations